Getting Started with Structural Equation Modeling
========================================================

## Jared E. Knowles
## jknowles@gmail.com


# Setting up your enviRonment

Getting started using structural equation modeling (SEM) in R can be daunting. There are 
lots of different packages for implementing SEM in R and there are different features 
of SEM that a user might be interested in implementing. A few packages you might come 
across can be found on the [CRAN Psychometrics Task View](http://cran.r-project.org/web/views/Psychometrics.html). 

For those who want to just dive in the `lavaan` package seems to offer the most 
comprehensive feature set for most SEM users and has a well thought out and easy 
to learn syntax for describing SEM models. To install lavaan, we just run:

```{r eval=FALSE, results='hide', echo=TRUE}
# Main version
install.packages("lavaan")

# Or to install the dev version
library(devtools)
install_github("lavaan", "yrosseel")
```

```{r setup, echo=FALSE, error=FALSE, message=FALSE, eval=TRUE, results='hide'}
opts_chunk$set(dev='svg', fig.width=3.75, fig.height=2.25, echo=TRUE, 
               message=FALSE, error=FALSE)

```

# Read in the data

Once we load up the lavaan package, we need to read in the dataset. `lavaan` accepts 
two different types of data, either a standard R dataframe, or a variance-covariance 
matrix. Since the latter is unfamiliar to us coming from the standard `lm` linear modeling 
framework in R, we'll start with reading in the simplest variance-covariance matrix 
possible and running a path analysis model. 

```{r matrixinput1}
library(lavaan)
mat1 <- matrix(c(1, 0, 0, 0.6, 1, 0, 0.33, 0.63, 1), 3, 3, byrow = TRUE)

colnames(mat1) <- rownames(mat1) <- c("ILL", "IMM", "DEP")

myN <- 500
print(mat1)
```

We could fit two models to this data:

1. DEP influences IMM influences ILL
2. IMM influences ILL influences DEP


```{r twomodels}

mod1 <- "ILL ~ IMM
        IMM ~ DEP"

mod1fit <- sem(mod1, sample.cov = mat1, sample.nobs = 500)

mod2 <- "DEP ~ ILL
             ILL ~ IMM"

mod2fit <- sem(mod2, sample.cov = mat1, sample.nobs = 500)

summary(mod1fit)
summary(mod2fit)

```


One of the best ways to understand an SEM model is to inspect the model visually 
using a path diagram. Luckily, this is easy to do in R:

```{r modelvis}
library(semPlot)
semPaths(mod1fit, what="est", layout="tree", title=TRUE, style="LISREL")
semPaths(mod2fit, what="est", layout="tree", title=TRUE, style="LISREL")
```

These two simple path models look great. 

# Appendix

```{r appendix}
citation(package="lavaan")
citation(package="semPlot")
     
sessionInfo()
     
```
