Getting Started with Structural Equation Modeling
========================================================

## Jared E. Knowles
## jknowles@gmail.com

# Introduction

For the analyst familiar with linear regression fitting structural equation models 
can at first feel strange. In the R environment, fitting structural equation models 
involves learning new modeling syntax, new plotting syntax, and often a new 
data input method. However, a quick reorientation and soon the user is exposed 
to the differences, fitting structural equation models can be a powerful tool in 
the analyst's toolkit. 

This tutorial will cover getting set up and running a few basic models using `lavaan` 
in R.[1] Future tutorials will cover:

* constructing latent variables
* comparing alternate models 
* multi-group analysis on larger datasets. 

# Setting up your enviRonment

Getting started using structural equation modeling (SEM) in R can be daunting. There are 
lots of different packages for implementing SEM in R and there are different features 
of SEM that a user might be interested in implementing. A few packages you might come 
across can be found on the [CRAN Psychometrics Task View](http://cran.r-project.org/web/views/Psychometrics.html). 

For those who want to just dive in the `lavaan` package seems to offer the most 
comprehensive feature set for most SEM users and has a well thought out and easy 
to learn syntax for describing SEM models. To install `lavaan`, we just run:

```{r eval=FALSE, results='hide', echo=TRUE}
# Main version
install.packages("lavaan")

# Or to install the dev version
library(devtools)
install_github("lavaan", "yrosseel")
```

```{r setup, echo=FALSE, error=FALSE, message=FALSE, eval=TRUE, results='hide'}
opts_chunk$set(dev='svg', fig.width=6, fig.height=6, echo=TRUE, 
               message=FALSE, error=FALSE)

```

# Read in the data

Once we load up the lavaan package, we need to read in the dataset. `lavaan` accepts 
two different types of data, either a standard R dataframe, or a variance-covariance 
matrix. Since the latter is unfamiliar to us coming from the standard `lm` linear modeling 
framework in R, we'll start with reading in the simplest variance-covariance matrix 
possible and running a path analysis model. 

```{r matrixinput1}
library(lavaan)
mat1 <- matrix(c(1, 0, 0, 0.6, 1, 0, 0.33, 0.63, 1), 3, 3, byrow = TRUE)

colnames(mat1) <- rownames(mat1) <- c("ILL", "IMM", "DEP")

myN <- 500
print(mat1)
# Note that we only input the lower triangle of the matrix. This is sufficient
# though we could put the whole matrix in if we like
```

Now we have a variance-covariance matrix in our environment named `mat1` and a 
variable `myN` corresponding to the number of observations in our dataset. Alternatively, 
we could provide R with the full dataset and it can derive `mat1` and `myN` itself. 

With this data we can construct two possible models: 

1. Depression (DEP) influences Immune System (IMM) influences Illness (ILL)
2. IMM influences ILL influences DEP

Using SEM we can evaluate which model best explains the covariances we observe in 
our data above. Fitting models in `lavaan` is a two step process. First, we create 
a text string that serves as the `lavaan` model and follows the `lavaan` [model 
syntax](http://www.inside-r.org/packages/cran/lavaan/docs/model.syntax). Next, we 
give `lavaan` the instructions on how to fit this model to the data using either the 
`cfa`, `lavaan`, or `sem` functions. Here we will use the `sem` function. Other 
functions will be covered in a future post. 

```{r twomodels}
# Specify the model 

mod1 <- "ILL ~ IMM 
        IMM ~ DEP"

# Give lavaan the command to fit the model 
mod1fit <- sem(mod1, sample.cov = mat1, sample.nobs = 500)

# Specify model 2

mod2 <- "DEP ~ ILL
        ILL ~ IMM"

mod2fit <- sem(mod2, sample.cov = mat1, sample.nobs = 500)
```

Now we have two objects stored in our environment for each model. We have the 
model string and the modelfit object. The model fit objects (`mod1fit` and `mod2fit`) 
are `lavaan` class objects. These are S4 objects with many supported methods, including 
the `summary` method which provides a lot of useful output: 

```{r modelsummaries}
# Summarize the model fit
summary(mod1fit)

summary(mod2fit)
```

One of the best ways to understand an SEM model is to inspect the model visually 
using a path diagram. Thanks to the `semPlot` package, 
this is easy to do in R.[2] First, install `semPlot`:

```{r eval=FALSE, results='hide', echo=TRUE}
# Official version
install.packages("semPlot")

# Or to install the dev version
library(devtools)
install_github("semPlot", "SachaEpskamp")
```

Next we load the library and make some path diagrams. 

```{r modelvis}
library(semPlot)
semPaths(mod1fit, what="est", layout="tree", title=TRUE, style="LISREL")
semPaths(mod2fit, what="est", layout="tree", title=TRUE, style="LISREL")
```

These two simple path models look great. But which is better? We can run a simple 
chi-square test on the `lavaan` objects `mod1fit` and `mod2fit`. 

```{r modelcomparison}
anova(mod1fit, mod2fit)
```

We can see that very clearly we prefer Model 2. Let's look at some properties of 
model 2 that we can access through the `lavaan` object with convenience functions. 

```{r modelexplore}
# Goodness of fit measures
fitMeasures(mod2fit)

# Estimates of the model parameters
parameterEstimates(mod2fit, ci=TRUE, boot.ci.type="norm")

# Modification indices
modindices(mod2fit, standardized=TRUE)
```

That's it. From inputing a variance-covariance matrix to fitting a model, 
drawing a path diagram, comparing to alternate models, and finally inspecting 
the parameters of the preferred model. `lavaan` is an amazing project which 
adds great capabilities to R. These will be explored in future posts. 

# Appendix

```{r appendix}
citation(package="lavaan")
citation(package="semPlot")
sessionInfo()
```


[1]: http://lavaan.ugent.be/ "The lavaan homepage"
[2]: http://sachaepskamp.com/semPlot/ "The semPlot homepage"